Network Data Model and BerlinMOD Benchmark

### 2010-07-19 Algebras/Network/BerlinMODNetwork/paper/PaperBerlinMODAndNetworkDataModel.tex
Widely finished but not read for correction.

The spatio-temporal minimum bounding box was introduced as an attribute to the
\dt{mgpoint} because the computation of this value is very expensive in the
NDM. Although each unit of an \dt{mgpoint} stays on the same route
at the same speed it may follow different spatial directions. For example, a
route may lead uphill in serpentine. A spatial bounding box only computed from
the spatial start and end position may not enclose all spatial positions of the
car within the unit. Therefore we always have to examine the spatial dimensions
of the $route\ interval$ passed within a unit to compute the units bounding box
using algorithm \ref{alg:unitbbox}.
\begin{algorithm}[H]
  \caption{\op{Berechnung Unit Bounding Box}}
  \label{alg:unitbbox}
  \begin{algorithmic}[1]
    \STATE Get route curve usint B-Tree index of $routes$ relation
    \STATE Extract subline of unit from route curve
    \STATE Compute bounding box of subline
    \STATE Add third dimension from unit time interval
  \end{algorithmic}
\end{algorithm}
If the number of routes in the $routes$ relation is $R$ the first step has a
time complexity of O($\log R$) time. If the route curve $r_i$ consists of $l_i$
line segments the time complexity of step 2 and 3 is O($l_i$) in the worst case.
Step 4 is done in O(1) time. Together we get a time complexity of
O($\log {R} + l_i$) to compute the bounding box of a single unit.

@@

Table \ref{tab:simplenetoperations} gives an overview of the simple benchmark
operations. The more complex operations are described in the sequel.
\begin{sidewaystable}
  \caption{Simple Operations on Network Data Types}
  \label{tab:simplenetoperations}
  %\begin{scriptsize}
  \begin{center}
  \begin{tabularx}{1.0\textwidth}{|l|l|X|X|}
  \hline
  \textbf{Name}& \textbf{Signature}& \textbf{Explanation}& \textbf{Complexity}\\
  \hline
\op{routes}& \dt{network} $\rightarrow$ \dt{relation} & Returns the $routes$
relation fo the \dt{network} object & O($R$)\\
  \hline
\op{no\_components}&\dt{mgpoint} $\rightarrow$ \dt{real}& Returns the number of
units of the \dt{mgpoint} &O(1)\\
  \hline
\op{length}&\dt{mgpoint} $\rightarrow$ \dt{real}& Returns the $length$ of the
argument.&O(1)\\
  \hline
\op{trajectory}&\dt{mgpoint} $\rightarrow$ \dt{gline}& Returns the $trajectory$
of the \dt{mgpoint} value as \dt{gline}&O($r$), if trajectory is defined, O($m \log r + r$) otherwise\\
  \hline
\op{units}&\dt{mgpoint} $\rightarrow$ \dt{stream}(\dt{ugpoint})& Returns a
stream
of the \dt{ugpoint} values of the \dt{mgpoint} value& O($m$)\\
  \hline
\op{initial}&\dt{mgpoint} $\rightarrow$ \dt{igpoint}& Returns the first position
and the start time of the \dt{mgpoint} value& O(1)\\
  \hline
\op{atinstant}&\dt{mgpoint} $\times$ \dt{instant} $\rightarrow$ \dt{igpoint}&
Returns the network position of the \dt{mgpoint} value at the given time instant&
O($\log m$)\\
  \hline
\op{val}&\dt{igpoint} $\rightarrow$ \dt{gpoint}& Returns the \dt{gpoint} of the
\dt{igpoint} value&O(1)\\
  \hline
\op{inst}&\dt{igpoint} $\rightarrow$ \dt{instant}& Returns the time instant of the
\dt{igpoint} value &O(1)\\
  \hline
\op{isempty}&\dt{gline} $\rightarrow$ \dt{bool}& Returns \true{} if the \dt{gline}
has no $route\ intervals$ & O(1)\\
  \hline
\op{routeintervals}&\dt{gline} $\rightarrow$ \dt{stream}(\dt{rectangle})& Returns
the $netbox$ for each $route\ interval$ of the \dt{gline} value& O($r$)\\
  \hline
\op{gpoint2rect}& \dt{gpoint} $\rightarrow$ \dt{rectangle} & Returns a $netbox$
for the \dt{gpoint} value & O(1)\\
\hline
\op{unitbox}& \dt{ugpoint} $\rightarrow$ \dt{rectangle3D} & Returns the $netbox$
of the \dt{ugpoint} & O(1)\\
  \hline
  \end{tabularx}
  \end{center}
  %\end{scriptsize}
\end{sidewaystable}

@@

In Appendix \ref{app:executableQueries} you can find the resulting executable
\secondo{} queries for NET in detail. The \secondo{} scripts with the queries
can be downloaded from our web page \cite{NetworkWeb}. In this Section we give
only a short overview over the \bmodb{} queries and their NET algorithms.

Every time we need a licence in the result or have a query licence number we need
an additional step in the TBA queries to connect the $dataMNtrip$ and $dataMcar$
relation using the $moid$ attribute and the corresponding B-Tree indexes.
We will not repeat this step at every single TBA query description.

Query 1 asks for the models of the cars with licence plate numbers from $QueryLicences$,
and query 2 for the number of vehicles that are ``passenger cars''. Both queries
deal only with standard attributes; so we only changed the relation names and the
B-Tree indexes to match the NET representation.

Query 3 searches for the positions of the ten cars from $QueryLicence1$ at the
ten time instants from $QueryInstants1$. We use the licence B-Tree to select the
ten cars and compute the positions of these ten cars for each of the ten time
instants from $QueryInstants1$ if the time instant is inside the definition time
of the trip.

Query 4 asks for the licence numbers of the cars that passed the points
from $QueryPointsNet$. We create a $netbox$ for each \dt{gpoint} in $QueryPointNet$
and use our specialised $netbox$ R-Tree of the \dt{mgpoint} \dt{RouteInterval} to
select the vehicles passing the given query points.

@@

Query 5 asks for the minimum distance between places where vehicles with
licences from $QueryLicence1$ and $QueryLicence2$ have been. We select the cars
with licence plate numbers from $QueryLicence1$ respectively $QueryLicences2$ using the
B-Tree over the $Licence$ attribute of the $dataSNcar$ relation. In the TBA, the
resulting trajectories for each car are aggregated into one single trajectory
for each car. In both approaches we create a \dt{line} value for each resulting
(aggregated) trajectory value of the \dt{mgpoint}s and compute the Euclidean distance
between these \dt{line} values for each pair of licences one from $QueryLicences1$ and one
from $QueryLicences2$.

Query 6 asks for the pairs of licences from ``trucks'' that have been as close as
10m or less to each other. We filter $dataSNcar$ relation, respectively $dataMcar$
relation to select the ``trucks'' and compute the spatio-temporal bounding box
of each trip of a ``truck''. We extend the spatial dimensions of the bounding boxes
by 5m in each spatial direction and retranslate the \dt{mgpoint} values into \dt{mpoint}
values in a first step. In a second step, we compute join the results from
step one with itself using the intersection of the bounding boxes as join criteria.
We filter the result to include all licence pairs of ``trucks'' that had sometimes
a distance lower than 10m. In the TBA, we additionally remove the duplicate licence
pairs from the result.

Query 7 asks for the licence plate numbers of the ``passenger'' cars that reached
the points from $QueryPoints$ first of all ``passenger'' cars during the
observation period. The first step to solve query 7 in the NDM is
equal to query 4. In a filter step we remove all ``not passenger'' cars from the
first intermediate result. We compute for each remaining candidate trip the times
the trip reaches first the query positions. We group the resulting time instants
by the identifiers of the query positions and compute the minimum time stamp of
each group, which is in fact the first time the query position was reached by a
car. In a last step the licences of the ``passenger'' cars reaching the query
positions at this first time instant are computed using the specialised
network-temporal index of the NDM.

@@

Query 10 asks when and where vehicles with licences from $QueryLicence1$ meet which
other vehicles (distance less than 3m). In the OBA we first retranslate every
\dt{mgpoint} value of $dataSNCar$ into a \dt{mpoint} value and extend the spatial
bounding box of each of this trips by 1.5 m in every spatial direction. After that we
select the ten candidate trips given by $QueryLicences1$, retranslate them and
extend their spatial bounding boxes in the same way. We join all trips from the
first two steps where the extended bounding boxes intersect and filter the
candidate pairs that have different licences and their distance is sometimes
less than 3m to each other. We compute the position of the \dt{mgpoint} at the
times the distance between the remaining candidate pairs of \dt{mpoint} was less
than 3 m and return the licence pairs and the network positions of the first car
when it has been closer than 3 m to the other one.

In the TBA we select the trips given by $QueryLicences1$ from
$dataMNtrip$, retranslate them into \dt{mpoint} values, and extend their spatio-temporal
bounding boxes by 3m in each spatial direction. After that we use the spatio-temporal
index of $dataMNtrip$ to select for each trip of the ten cars, the cars
of $dataMNtrip$ which spatio-temporal bounding boxes intersect the extended spatio-temporal
bounding boxes built before. For every pair of candidate trips we
retranslate the second trip and use the Euclidean Distance function for \dt{mpoint}
values to determine the times when the both \dt{mgpoint} had a distance less than 3m.
At last we restrict the trip of the query \dt{mgpoint} to this times and
aggregate the resulting trips into one single trip for each licence pair.

In our experiments we tried out several indexes to support a faster query execution
of query 10 including the MON-Tree presented in \cite{MONTreeAlmeidaGeoinformatica}. The MON-Tree showed
very good CPU times but never the less the total run time was very high. In the end
the simple form described above showed the best complete run time performance of all
indexes.

@@

Query 15 asks for the vehicles passing a point from $QueryPoints1Net$ during a period
from $QueryPeriods1$. Analogous to query 14 we build temporal-netboxes of the
of the query parameters to select the candidate trips using the temporal-network
position index and refine the result filtering the candidates really fullfilling
the query constraints.

Query 16 asks for the licence pairs one from $QueryLicence1$ and one from $QueryLicence2$
of vehicles, which were both present in a region from $QueryRegions1Net$ within a period
from $QueryPeriods1$, but did not meet there and then. We select the candidate trips
using the licence B-Tree of $dataSNcar$ relation and restrict the resulting trips to be
\op{present} during the query periods and \op{inside} the query region. This is done
one time for the licences from $QueryLicences1$ and one time for the licences from
$QueryLicences2$. The both intermediate results are joined and filtered to get the
trips of different cars which where at the same period in the same region
without meeting each other there and then. In the TBA we have
to do a additional selection from trips with the $moids$ belonging to the cars
selected before by the licences and remove duplicates of licence pairs from the
same period and region.

Query 17 asks for the points from $QueryPointsNet$ that have been visited by a
maximum number of different vehicles. In a first step we use almost the query
algorithm from query 4 to select the trips passing a given query
point. After that we group the cars passing query points by the ids of the
query points and count the number of cars passing this query point. In a last
step the point(s) with the maximum number of passing cars is(are) selected.
In the TBA we have to remove duplicate vehicles from the result list before we
count the number of passing cars.
\section{Experimental Setup}
\label{sec:scenario}
For our experiments we used a standard personal computer with an AMD Phenom II X4
Quad Core 2.95 GHz CPU, 8 GB main memory, and 2 TB hard disk. We installed the
Linux openSUSE 11.2 as operating system, \secondo{} DBMS version 3.0, and
the \bmodb{} version provided in the web \cite{NetworkWeb}.

We generated three databases with different amounts of data using the data generation
script of the \bmodb{} with the $scalefactor$ 0.05, 0.2, and 1.0. The following
steps are done with all three databases. We first created the \bmodb{} data and indexes
using the script ``BerlinMOD\_CreateObjects.SEC'' for SPACE. The NET
representation of the databases was generated by the the script ``Network\_CreateObjects.SEC''
that uses the algorithms and builds the indexes described in Section \ref{sec:Translation}.

Table \ref{tab:dbstat} shows the created amounts of data for the different $scalefactor$
values in both data models. As you can see, NET NDM needs less
than 40\% of the storage space of the \bmodb{} data model. The main cause is that
the same trip is represented by less than 50\% of the units in NET
compared to SPACE. This is a very good result and we expect this effect to
increase if the cars make long distance
trips instead of moving in a single town like they do in the benchmark. In towns
cars more often change the street or the velocity than cars that do long distance
trips and so the compact route representation in NET should become
more effective than in the town.
\begin{table}[H]
\begin{center}
\begin{scriptsize}
\begin{tabularx}{1.0\textwidth}{|X|c|c|c|c|c|c|}
\hline
&\multicolumn{2}{c|}{\textbf{Scalefactor 0.05}}&\multicolumn{2}{c|}{\textbf{Scalefactor 0.2}}&\multicolumn{2}{c|}{\textbf{Scalefactor 1.0}}\\
\hline
Number of Cars&\multicolumn{2}{c|}{447}&\multicolumn{2}{c|}{894}&\multicolumn{2}{c|}{2000}\\
\hline
Number of Days&\multicolumn{2}{c|}{6}&\multicolumn{2}{c|}{13}&\multicolumn{2}{c|}{28}\\
\hline
Data Generation&\multicolumn{2}{c|}{164.761s}&\multicolumn{2}{c|}{587.299s}&\multicolumn{2}{c|}{3177.46s}\\
\hline
&SPACE&NET&SPACE&NET&SPACE&NET\\
\hline
Data Translation
and Index Build&301.72s&535.65s&1,362.72s&2,190.45s&7,419.13s&11,144.13s\\
\hline
Number of Units&2,646,026&1,260,888&11,296,682&5,346,971&52,140,685&24,697,709\\
\hline
Total Storage Space&2.26 GB&0.86 GB&9.51 GB&3.69 GB&45,76 GB&17.28 GB\\
Data&0.79 GB&0.44 GB&3.35 GB&1.83 GB&15.47 GB& 8.40 GB\\
Indexes&1.48 GB&0.42 GB&6.16 GB&1.86 GB&30.30 GB&8.89 GB\\
\hline
\end{tabularx}
\end{scriptsize}
\caption{Database Statistics}
\label{tab:dbstat}
\end{center}
\end{table}
The long creation time of the NET representation is
caused by the expensive mapping of spatial and spatio-temporal positions into
network positions. The indexes themselves are built faster in NET representation
than in SPACE representation because they have less entries.

@@

that the source data of the street map of the \bmodb{} is not well defined in all places.
Figure \ref{fig:routefailure} shows two examples for the street map failures. Using

@@

approaches. The tables in Figure \ref{fig:compruntimes} and the graphic in

@@

50\% less than the total query run time of SPACE at each scale factor.

@@

Although the query algorithms for both data models and approaches are almost the same,
SPACE outperforms NET for query 3 by tenths of a second. On the one hand, the
number of units in NET is less than the number of units in the SPACE,
such that the unit which contains the query time instant should be found faster.
But on the other hand, a \dt{gpoint} value has more internal elements (3 \dt{int},
1 \dt{real}, and 1 \dt{bool}) than a \dt{point} value (2 \dt{real}, and 1 \dt{bool}),
such that each unit in NET is much bigger than in SPACE, therefore computing and
copying the result needs a little more time in NET than in SPACE.

@@

After the very good results from query 4 we did not expect query 7 to have such
bad results in the run time comparison. In fact SPACE outperforms NET significantly
in both approaches and at all $scalefactors$. in TBA, while at the smaller scale factors in
OBA the NDM outperforms the DMFS. We think there are two main causes: On the one hand we
have to do the expensive operation \dt{at} for \dt{mgpoint} for the double number
of query \dt{gpoint} compared with the DMFS and
on the other hand the test points out a weakness of the NDM
implementation of the operation \op{at} for \dt{mgpoint}. But in the end, NDM looses
at scale factor 1.0 less than 45 seconds in the OBA respectively less
than 120 seconds in TBA, what is not much compared with the advantages in the
other benchmark queries.

Query 8 is a very fast query in both data models, although the query run time of
the NDM is more than 50\% less than the query run time of the
DMFS. This is caused by the
$length$ attribute of the \dt{mgpoint} and the smaller number of units of a
\dt{mgpoint} compared with the corresponding \dt{mpoint}.

For query 9, the NDM outperforms the DMFS by orders of magnitude. The advantages
named in the analysis of query 8's run time results have a much higher impact
when the number of examined trips becomes bigger. At scale factor 1.0 this saves
more than 10 min time in the OBA and more than 50 min time in the TBA.

The ratio of the run times of query 10 changes between the amounts of data and
both data models. In the OBA and at scale factor 1.0 in the TBA the NDM outperforms
the DMFS at all scale factors, while in the TBA at the two small databases the
DMFS outperforms the NDM significantly. Before our experiments we expected that
the DMFS would outperform the NDM in all cases, because of the expensive
retranslation of intermediate results. So why is the NDM faster ($>$ 20 min in
OBA and $>$ 3 min in TBA at scale factor 1.0) than the DMFS?
In the OBA we use bounding boxes for a preselect of candidate trips that step is not
performed in the DMFS. In the TBA the results are only better for the big amounts
of data we think this is due to the fact that the number of units in \dt{mgpoint}
values is always smaller than in \dt{mpoint} values such that the final aggregation
of the different trips of the same cars can be done faster in the NDM
than in the DMFS.

Query 11 is identical with the first part of query 12. So it is surprising that
the run time of query 11 at scalefactor 1.0 is longer than the run time of query 12,
which does additional computations. In our experiments with the different queries
we have seen that there exist numerous cache effects depending on the sequence of
the queries. So we think that query 12 takes profit cache effects resulting from
query 11 running immediately before query 12. Another weakness of the NDM pointed
out by the run times of query 11 and query 14 is that our
network-temporal position index has bad run times for query $netbox$ objects
constructed from a single \dt{gpoint} and a single time instant. This becomes worse
with a higher number of indexed units. As you can see at query 15 this does not
hold for query $netboxes$ constructed from a single \dt{gpoint} and a time interval.
We have to spend some more work to figure out the problem and develop a better
network-temporal position index to improve our NDM system.

In our experiments we also tested the MON-Tree \cite{MONTreeAlmeidaGeoinformatica} as network-temporal
index but the elapsed run time performance was not good, although the CPU run times
was were small.

The bad performance of the network-temporal position index is also shown by query 13.
The NDM outperforms the DMFS significantly, but we do not use any index in the
executable NDM queries, while the DMFS uses its spatio-temporal index to preselect
candidate trips. The same holds for query 17.

The NDM version of query 16 takes profit from the smaller number
of units in the NDM and outperforms the data model of free movement
in two dimensional space.

>>>>

The spatio-temporal minimum bounding box was introduced as an attribute to the
\dt{mgpoint} because the computation of this value is very expensive in the
NDM. Although each unit of an \dt{mgpoint} stays on the same route
at the same speed it may follow different spatial directions. For example, a
route may lead uphill in serpentine. A spatial bounding box only computed from
the spatial start and end position may not enclose all spatial positions of the
car within the unit. Therefore we always have to examine the spatial dimensions
of the $route\ interval$ passed within a unit to compute the units bounding box
using algorithm \ref{alg:unitbbox}.
\begin{algorithm}[H]
  \caption{Berechnung Unit Bounding Box}
  \label{alg:unitbbox}
  \begin{algorithmic}[1]
    \STATE Get route curve usint B-Tree index of $routes$ relation
    \STATE Extract subline of unit from route curve
    \STATE Compute bounding box of subline
    \STATE Add third dimension from unit time interval
  \end{algorithmic}
\end{algorithm}
If the number of routes in the $routes$ relation is $R$ the first step has a
time complexity of O($\log R$) time. If the route curve $r_i$ consists of $l_i$
line segments the time complexity of step 2 and 3 is O($l_i$) in the worst case.
Step 4 is done in O(1) time. Together we get a time complexity of
O($\log {R} + l_i$) to compute the bounding box of a single unit.

@@

Table \ref{tab:simplenetoperations} gives an overview of the simple benchmark
operations. The more complex operations are described in the sequel.
\begin{sidewaystable}
  \caption{Simple Operations on Network Data Types}
  \label{tab:simplenetoperations}
  %\begin{scriptsize}
  \begin{center}
  \begin{tabularx}{1.0\textwidth}{|l|l|X|X|}
  \hline
  \textbf{Name}& \textbf{Signature}& \textbf{Explanation}& \textbf{Complexity}\\
  \hline
\op{routes}& \dt{network} $\rightarrow$ \dt{relation} & Returns the $routes$
relation fo the \dt{network} object & O($R$)\\
  \hline
\op{no\_components}&\dt{mgpoint} $\rightarrow$ \dt{real}& Returns the number of
units of the \dt{mgpoint} &O(1)\\
  \hline
\op{length}&\dt{mgpoint} $\rightarrow$ \dt{real}& Returns the $length$ of the
argument.&O(1)\\
  \hline
\op{trajectory}&\dt{mgpoint} $\rightarrow$ \dt{gline}& Returns the $trajectory$
of the \dt{mgpoint} value as \dt{gline}&O($r$), if trajectory is defined,\\
&&&O($m \log r + r$) otherwise\\
  \hline
\op{units}&\dt{mgpoint} $\rightarrow$ \dt{stream}(\dt{ugpoint})& Returns a
stream
of the \dt{ugpoint} values of the \dt{mgpoint} value& O($m$)\\
  \hline
\op{initial}&\dt{mgpoint} $\rightarrow$ \dt{igpoint}& Returns the first position
and the start time of the \dt{mgpoint} value& O(1)\\
  \hline
\op{atinstant}&\dt{mgpoint} $\times$ \dt{instant} $\rightarrow$ \dt{igpoint}&
Returns the network position of the \dt{mgpoint} value at the given time instant&
O($\log m$)\\
  \hline
\op{val}&\dt{igpoint} $\rightarrow$ \dt{gpoint}& Returns the \dt{gpoint} of the
\dt{igpoint} value&O(1)\\
  \hline
\op{inst}&\dt{igpoint} $\rightarrow$ \dt{instant}& Returns the time instant of the
\dt{igpoint} value &O(1)\\
  \hline
\op{isempty}&\dt{gline} $\rightarrow$ \dt{bool}& Returns \true{} if the \dt{gline}
has no $route\ intervals$ & O(1)\\
  \hline
\op{routeintervals}&\dt{gline} $\rightarrow$ \dt{stream}(\dt{rectangle})& Returns
the $netbox$ for each $route\ interval$ of the \dt{gline} value& O($r$)\\
  \hline
\op{gpoint2rect}& \dt{gpoint} $\rightarrow$ \dt{rectangle} & Returns a $netbox$
for the \dt{gpoint} value & O(1)\\
\hline
\op{unitbox}& \dt{ugpoint} $\rightarrow$ \dt{rectangle3D} & Returns the $netbox$
of the \dt{ugpoint} & O(1)\\
  \hline
  \end{tabularx}
  \end{center}
  %\end{scriptsize}
\end{sidewaystable}

@@

In Appendix \ref{app:executableQueries} the resulting executable
\secondo{} queries are for NET are given in detail. The \secondo{} scripts with
the queries can be downloaded from our web page \cite{NetworkWeb}.
In this Section we give a short overview of the query algorithms used for the
NET representation of the \bmodb{}.

Every time we need a licence in the result or have a query licence number we need
an additional step in the TBA queries to connect the $dataMNtrip$ and $dataMcar$
relation using the $moid$ attribute and the corresponding B-Tree indexes
$dataMNtrip\_Moid\_btree$ respectively $dataMcar\_Moid\_btree$.
We will not repeat this step at every single TBA query description.

Query 1 asks for the models of the cars with licence plate numbers from $QueryLicences$,
and query 2 for the number of vehicles that are ``passenger cars''. Both queries
deal only with standard attributes; so we only changed the relation names and the
B-Tree indexes to match the NET representation of the \bmodb{}.

Query 3 searches for the positions of the ten cars from $QueryLicence1$ at the
ten time instants from $QueryInstants1$. We use the licence B-Tree to select the
ten cars and compute the positions of these ten cars for each of the ten time
instants from $QueryInstants1$, if the time instant is inside the definition time
of the trip.

Query 4 asks for the licence numbers of the cars that passed the points
from $QueryPointsNet$. We create a $netbox$ for each \dt{gpoint} in $QueryPointNet$
and use our specialised R-Tree $dataSNcar\_TrajBoxNet$ respectively
$dataMNtrip\_TrajBoxNet$ to select the vehicles passing the given query points.

@@

Query 5 asks for the minimum distance between places where vehicles with
licences from $QueryLicence1$ and $QueryLicence2$ have been. We select the cars
with licence plate numbers from $QueryLicence1$ respectively $QueryLicences2$ using the
B-Tree over the $Licence$ attribute. In the TBA, the
resulting trajectories for each car are aggregated into one single $trajectory$
for each car. In both approaches we create a \dt{line} value for each resulting
(aggregated) $trajectory$ of the \dt{mgpoint}s and compute the Euclidean distance
between these \dt{line} values for each pair of licences one from $QueryLicences1$
and one from $QueryLicences2$.

Query 6 asks for the pairs of licences from ``trucks'' that have been as close as
10m or less to each other. We select the ``trucks'' from $dataSNcar$ relation,
respectively $dataMcar$ relation, extend their spatio-temporal
bounding boxes in all spatial dimensions by 5m, and retranslate the
\dt{mgpoint} values into \dt{mpoint} values in a first step. In a second step,
we join the results from step one with itself using the intersection of the bounding boxes as join criteria. We filter the result to include all licence pairs of
``trucks'' that had sometimes a distance lower than 10m and remove the duplicate
licence pairs from the result.

Query 7 asks for the licence plate numbers of the ``passenger'' cars that reached
the points from $QueryPointsNet$ first of all ``passenger'' cars during the
observation period. We select all ``passenger'' cars that passed one
of the points from $QueryPointsNet$ before we compute for each remaining candidate
trip the times the trip reaches first the query positions.
We group the resulting time instants
by the identifiers of the query positions and compute the minimum time stamp of
each group, which is in fact the first time the query position was reached by a
car. In a last step the licences of the ``passenger'' cars reaching the query
positions at this first time instant are selected.

@@

Query 10 asks when and where vehicles with licences from $QueryLicence1$ meet which
other vehicles (distance less than 3m). In the OBA we first retranslate every
\dt{mgpoint} value of $dataSNCar$ into a \dt{mpoint} value and extend the spatial
bounding box of each of this trips by 1.5 m in every spatial direction. After that we
select the ten candidate trips given by $QueryLicences1$, retranslate them and
extend their spatial bounding boxes in the same way. We join all trips from the
first two steps where the extended bounding boxes intersect and filter the
candidate pairs that have different licences and sometimes a distance less than
3m to each other. We compute the position of the \dt{mgpoint} at the
times the distance was less than 3 m and return the licence pairs and the network positions of the first car when it has been closer than 3 m to the other one.

In the TBA, we select the trips given by $QueryLicences1$ from
$dataMNtrip$, retranslate them into \dt{mpoint} values, and extend their spatio-temporal
bounding boxes by 3m in each spatial direction. After that, we use the spatio-temporal
index of $dataMNtrip$ to select the cars of $dataMNtrip$ with intersecting
spatio-temporal bounding boxes. For every pair of candidate trips we
retranslate the second trip and use the Euclidean Distance function for \dt{mpoint}
values to determine the times when the both \dt{mgpoint} had a distance less than 3m.
At last we restrict the trip of the query \dt{mgpoint} to this times and
aggregate the resulting trips into one single trip for each licence pair.

In our experiments we tried out several indexes to support a faster query execution
of query 10 including the MON-Tree presented in \cite{MONTreeAlmeidaGeoinformatica}.
The MON-Tree used very small amounts of CPU time, but nevertheless, the total query
run time was very high if we used the MON-Tree. In the end the simple form described
above showed the best complete run time performance of all indexes.

@@

Query 15 asks for the vehicles passing a point from $QueryPoints1Net$ during a
period from $QueryPeriods1$. Almost similar to query 14 we select the candidate
trips building temporal-netboxes of the query parameters and using TNPI. The
result is refined by a filter step selecting the candidates really fullfilling
the query constraints.

Query 16 asks for the licence pairs one from $QueryLicence1$ and one from
$QueryLicence2$ of vehicles, which were both present in a region from
$QueryRegions1Net$ within a period from $QueryPeriods1$, but did not meet there
and then. We select the candidate trips using the licence B-Tree and restrict
the resulting trips to be \op{present} during the query periods and move
\op{inside} the query regions. This is done one time for the licences from $QueryLicences1$ and one time for the licences from $QueryLicences2$.
The both intermediate results are joined and filtered to get the
trips of different cars which where at the same period in the same region
without meeting each other there and then. In the TBA we have
to remove duplicates of licence pairs from the same period and region.

Query 17 asks for the points from $QueryPointsNet$ that have been visited by a
maximum number of different vehicles. In a first step, we use almost the query
algorithm from query 4 to select the trips passing a given query
point. After that we group the cars passing query points by the ids of the
query points and count the number of cars passing this query point. In a last
step the point / points with the maximum number of passing cars is / are selected.
In the TBA we have to remove duplicate vehicles from the result list before we
count the number of passing cars.
\section{Experimental Setup}
\label{sec:scenario}
For our experiments we used a standard personal computer with an AMD Phenom II X4
Quad Core 2.95 GHz CPU, 8 GB main memory, and 2 TB hard disk. We installed the
Linux openSUSE 11.2 as operating system, \secondo{} DBMS version 3.0, and
the \bmodb{} version provided in the web \cite{NetworkWeb}.

We generated three databases with different amounts of data using the data generation
script of the \bmodb{} with the $scalefactor$ 0.05, 0.2, and 1.0. The following
steps are done with all three databases. We first created the \bmodb{} data and
indexes using the script ``BerlinMOD\_CreateObjects.SEC'' for SPACE,
and the script ``Network\_CreateObjects.SEC'' for NET representation. The latter
script uses the algorithms and builds the indexes described in Section
 \ref{sec:Translation}.

Table \ref{tab:dbstat} shows the created amounts of data for the different
$scalefactor$ values in both data models. As you can see, NET needs less
than 40\% of the storage space of the \bmodb{} data model. The main cause is that
the same trip is represented by less than 50\% of the units in NET
compared to SPACE. This is a very good result and we expect this effect to
increase if the cars make long distance
trips instead of moving in a single town like they do in the benchmark. In towns
cars more often change the street or the velocity than cars that do long distance
trips and so the compact route representation in NET should become
more effective than in the town.
\begin{table}[H]
\begin{center}
\begin{scriptsize}
\begin{tabularx}{1.0\textwidth}{|X|c|c|c|c|c|c|}
\hline
&\multicolumn{2}{c|}{\textbf{Scalefactor 0.05}}&\multicolumn{2}{c|}{\textbf{Scalefactor 0.2}}&\multicolumn{2}{c|}{\textbf{Scalefactor 1.0}}\\
\hline
Number of Cars&\multicolumn{2}{c|}{447}&\multicolumn{2}{c|}{894}&\multicolumn{2}{c|}{2000}\\
\hline
Number of Days&\multicolumn{2}{c|}{6}&\multicolumn{2}{c|}{13}&\multicolumn{2}{c|}{28}\\
\hline
Data Generation&\multicolumn{2}{c|}{164.761s}&\multicolumn{2}{c|}{587.299s}&\multicolumn{2}{c|}{3177.46s}\\
\hline
&SPACE&NET&SPACE&NET&SPACE&NET\\
\hline
Data Translation
and Index Build&301.72s&535.65s&1,362.72s&2,190.45s&7,419.13s&11,144.13s\\
\hline
Number of Units&2,646,026&1,260,888&11,296,682&5,346,971&52,140,685&24,697,709\\
\hline
Total Storage Space&2.26 GB&0.86 GB&9.51 GB&3.69 GB&45,76 GB&17.28 GB\\
Data&0.79 GB&0.44 GB&3.35 GB&1.83 GB&15.47 GB& 8.40 GB\\
Indexes&1.48 GB&0.42 GB&6.16 GB&1.86 GB&30.30 GB&8.89 GB\\
\hline
\end{tabularx}
\end{scriptsize}
\caption{Database Statistics}
\label{tab:dbstat}
\end{center}
\end{table}
The long creation time of the NET representation is
caused by the expensive mapping of spatial and spatio-temporal positions into
network positions. The indexes themselves are built faster in NET representation
than in SPACE representation because they have less entries.

@@

that the source data of the street map of the \bmodb{} is not well defined in all
places. Figure \ref{fig:routefailure} shows two examples for the street map
failures. Using

@@

approaches. The tables in Figure \ref{fig:compruntimes} and the graphic in Figure

@@

50\% less than the total query run time of SPACE at each $scalefactor$.

@@

Although the query algorithms for both data models and approaches are almost the
same, SPACE outperforms NET for query 3 by tenths of a second. On the one hand, the
number of units in NET is less than the number of units in SPACE,
such that the unit which contains the query time instant should be found faster.
But on the other hand, a \dt{gpoint} value has more internal elements (3 \dt{int},
1 \dt{real}, and 1 \dt{bool}) than a \dt{point} value (2 \dt{real}, and 1 \dt{bool}),
such that each unit in NET has much four more elements than in SPACE, therefore
access and copying needs a little more time in NET than in SPACE.

@@

After the very good results from query 4 we did not expect SPACE to outperform
NET at query 7 in almost all cases. Only at $scalfactor$ 1.0 in OBA NET outperforms
SPACE. The weakness of NET is that we have to do the expensive \op{at} operation
for the double number of query \dt{gpoint} compared with SPACE
(see Section \ref{sec:translateSTdata}).
But in the end, NET looses at $scalefactor$ 1.0 around 1 minute, what is not much
compared with the advantages in the other benchmark queries.

Query 8 is a very fast query in both data models, although the query run time of
NET is more than 50\% less than the query run time of SPACE. This is caused by the
$length$ attribute of the \dt{mgpoint} and the smaller number of units of a
\dt{mgpoint} compared with the corresponding \dt{mpoint}.

For query 9, NET outperforms SPACE by orders of magnitude. The advantages
named in the analysis of query 8 have a much higher impact
when the number of examined trips becomes bigger. At $scalefactor$ 1.0 this saves
more than 40 minutes time in the OBA and more than 55 minutes time in the TBA.

The ratio of the run times of query 10 changes between the amounts of data and
both data models. In the OBA NET outperforms SPACE at all $scalefactors$,
while in the TBA SPACE outperforms NET significantly.
Before our experiments we expected that SPACE would outperform NET in all cases,
because of the expensive retranslation of intermediate results.
So why is NET 80 minutes faster at $scalefactor$ 1.0 in OBA than SPACE?
In NET we use bounding boxes to preselect candidate trips that step is not
performed in SPACE. The disadvantage of NET in TBA becomes smaller at higher
$scalefactors$ we think this is due to the fact that the number of units in
\dt{mgpoint} values is always smaller than in \dt{mpoint} values such that the
final aggregation of the different trips of the same cars can be done faster
in NET than SPACE.

In our experiments we also tested the MON-Tree \cite{MONTreeAlmeidaGeoinformatica}
as spatio-temporal index. But, although the CPU run time was very small, the
elapsed run time performance was very bad. Such that the primitive index outperformed
the MON-Tree in all cases.

Query 11 is identical with the first part of query 12. So it is surprising that
the run time of query 11 at $scalefactor$ 1.0 is longer than the run time of query 12,
which does additional computations. In our experiments with the different queries
we have seen that there exist numerous cache effects depending on the sequence of
the queries. So we think that query 12 takes profit of cache effects resulting from
query 11 running immediately before query 12. Another weakness of NET pointed
out by the run times of query 11 and some other queries is that our TNPI has bad
run times for query $netboxes$ created from a single \dt{gpoint} and a single
time instant. This becomes worse with a higher number of indexed units. As you can
see at query 15 this does not hold for query $netboxes$ constructed from a single
\dt{gpoint} and a time interval. We have to spend some more work to figure out
the problem and develop a better TNPI to improve our NET implementation.

The bad performance of the network-temporal position index is also shown by query 13.
NET outperforms SPACE significantly, but we do not use any index in the
executable NET queries, while SPACE uses its spatio-temporal index to preselect
candidate trips. The same holds for query 17.

The NET version of query 16 takes profit from the smaller number
of units in NET and outperforms the data model of free movement
in two dimensional space.
